---
title: "02-tidycensus"
output: html_document
date: "2025-03-25"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Tidycensus introduction

## Decenial census

```{r}
library(tidycensus)
#census_api_key("2ae6b4b06c6b461ac27efd364c2fa133d96c0051", install = T)
```

```{r}
api_key <- Sys.getenv("CENSUS_API_KEY")
```

```{r}
total_population_10 <- get_decennial(
  geography = "state", 
  variables = "P001001", 
  year = 2010
)
```

```{r}
total_population_10
```

## Decenial summary

2020 Decennial Census data are available from the PL 94-171 Redistricting summary file, which is specified with `sumfile = "pl"` and is also available for 2010. The Redistricting summary files include a limited subset of variables from the decennial US Census to be used for legislative redistricting. These variables include total population and housing units; race and ethnicity; voting-age population; and group quarters population. For example, the code below retrieves information on the American Indian & Alaska Native population by state from the 2020 decennial Census.

The argument `sumfile = "pl"` is assumed (and in turn not required) when users request data for 2020 and will remain so until the main Demographic and Housing Characteristics File is released in mid-to-late 2022

```{r}
aian <- get_decennial(
  geography = "state", 
  variables = "P1_005N", 
  year = 2020, 
  sumfile = "pl"
)
aian
```

## American Community Survey

The example below fetches data on the number of residents born in Mexico by state.

```{r}
born_in_mexico <- get_acs(
  geography = "state", 
  variables = "B05006_150",
  year = 2020
)
born_in_mexico
```

`survey` defaults to the 5-year ACS; however this can be changed to the 1-year ACS by using the argument `survey = "acs1"`. For example, the following code will fetch data from the 1-year ACS for 2019:

```{r}
born_in_mexico_1yr <- get_acs(
  geography = "state", 
  variables = "B05006_150", 
  survey = "acs1", 
  year = 2019
)
born_in_mexico_1yr
```

Variables from the ACS detailed tables, data profiles, summary tables, comparison profile, and supplemental estimates are available through **tidycensus**'s [`get_acs()`](https://walker-data.com/tidycensus/reference/get_acs.html) function; the function will auto-detect from which dataset to look for variables based on their names. Alternatively, users can supply a table name to the table parameter in get_acs(); this will return data for every variable in that table. For example, to get all variables associated with table B01001, which covers sex broken down by age, from the 2016-2020 5-year ACS:

```{r}
age_table <- get_acs(
  geography = "state", 
  table = "B01001", 
  year = 2020, 
  cache_table = T
)
age_table
```

## Geography and variables

For core-based statistical areas and zip code tabulation areas, two heavily-requested geographies, the aliases `"cbsa"` and `"zcta"` can be used, respectively, to fetch data for those geographies.

```{r}
cbsa_population <- get_acs(
  geography = "cbsa", 
  variables = "B01003_001", 
  year = 2020
)
cbsa_population
```

## Geographic subsets

```{r}
nm_income <- get_acs(
  geography = "county", 
  variables = "B19013_001", 
  state = "NM", 
  year = 2020
)
nm_income
```

Smaller geographies like Census tracts can also be subsetted by county. Given that Census tracts nest neatly within counties (and do not cross county boundaries), we can request all Census tracts for a given county by using the optional `county` parameter.

```{r}
bernalillo_income <- get_acs(
  geography = "tract", 
  variables = "B19013_001", 
  state = "NM", 
  county = "Bernalillo", 
  year = 2020
)
bernalillo_income
```

```{r}
nrow(nm_income)
```

```{r}
nm_income_1yr <- get_acs(
  geography = "county", 
  variables = "B19013_001", 
  state = "NM", 
  year = 2019, 
  survey = "acs1"
)
nrow(nm_income_1yr)
nm_income_1yr
```

## Searching for variables

> [`load_variables()`](https://walker-data.com/tidycensus/reference/load_variables.html)

```{r}
v16 <- load_variables(2016, "acs5", cache = T)
v16
```

```{r}
View(v16)
```

By default, **tidycensus** returns a tibble of ACS or decennial Census data in "tidy" format. For decennial Census data, this will include four columns:

-   `GEOID`, representing the Census ID code that uniquely identifies the geographic unit;
-   `NAME`, which represents a descriptive name of the unit;
-   `variable`, which contains information on the Census variable name corresponding to that row;
-   `value`, which contains the data values for each unit-variable combination. For ACS data, two columns replace the `value` column: `estimate`, which represents the ACS estimate, and `moe`, representing the margin of error around that estimate.

## Data structures

```{r}
hhinc <- get_acs(
  geography = "state", 
  table = "B19001", 
  survey = "acs1",
  year = 2016
)
hhinc
```

```{r}
hhinc_wide <- get_acs(
  geography = "state", 
  table = "B19001", 
  survey = "acs1",
  year = 2016, 
  output = "wide"
)
hhinc_wide
```

## GEOIDs

```{r}
bernalillo_blocks <- get_decennial(
  geography = "block", 
  variables = "H1_001N",
  state = "NM", 
  county = "Bernalillo", 
  year = 2020, 
  sumfile = "pl"
)
bernalillo_blocks
```

The GEOID value breaks down as follows:

-   The first two digits, **40**, correspond to the [Federal Information Processing Series (FIPS) code](https://www.census.gov/library/reference/code-lists/ansi.html) for the state of Oklahoma. All states and US territories, along with other geographies at which the Census Bureau tabulates data, will have a FIPS code that can uniquely identify that geography.
-   Digits 3 through 5, **025**, are representative of Cimarron County. These three digits will uniquely identify Cimarron County within Oklahoma. County codes are generally combined with their corresponding state codes to uniquely identify a county within the United States, as three-digit codes will be repeated across states. Cimarron County's code in this example would be **40025**.
-   The next six digits, **950300**, represent the block's Census tract. The tract name in the `NAME` column is Census Tract 9503; the six-digit tract ID is right-padded with zeroes.
-   The twelfth digit, **1**, represents the parent block group of the Census block. As there are no more than nine block groups in any Census tract, the block group name will not exceed 9.
-   The last three digits, **110**, represent the individual Census block, though these digits are combined with the parent block group digit to form the block's name.

## Renaming variables

ables = c(medinc = "B19013_001",

medage = "B01002_001"),

year = 2020

)

nm

```{r}
nm <- get_acs(
  geography = "county", 
  state = "New Mexico", 
  variables = c(medinc = "B19013_001", 
                medage = "B01002_001"), 
  year = 2020
)
nm
```

```{r}
nm_wide <- get_acs(
  geography = "county", 
  state = "New Mexico", 
  variables = c(medinc = "B19013_001", 
                medage = "B01002_001"), 
  output = "wide",
  year = 2020
)
nm_wide
```

## Get estimates

One advantage of using the PEP to retrieve data is that allows you to access the indicators used to produce the intercensal population estimates. These indicators can be specified as variables direction in the [`get_estimates()`](https://walker-data.com/tidycensus/reference/get_estimates.html) function in **tidycensus**, or requested in bulk by using the `product` argument. The products available include `"population"`, `"components"`, `"housing"`, and `"characteristics"`. For example, we can request all components of change population estimates for 2019 for a specific county

```{r}
library(tidyverse)

bernalillo_components <- get_estimates(
  geography = "county", 
  product = "components", 
  state = "NM", 
  county = "Bernalillo", 
  year = 2019
)
bernalillo_components
```

The `product = "characteristics"` argument also has some unique options. The argument `breakdown` lets users get breakdowns of population estimates for the US, states, and counties by `"AGEGROUP"`, `"RACE"`, `"SEX"`, or `"HISP"` (Hispanic origin). If set to `TRUE`, the `breakdown_labels` argument will return informative labels for the population estimates. For example, to get population estimates by sex and Hispanic origin for metropolitan areas, we can use the following code:

```{r}
nm_sex_hisp <- get_estimates(
  geography = "state", 
  product = "characteristics", 
  breakdown = c("SEX", "HISP"), 
  breakdown_labels = T,
  state = "NM", 
  year = 2019
)
nm_sex_hisp
```

## Get flows

```{r}
bernalillo_migration <- get_flows(
  geography = "county", 
  state = "NM", 
  county = "Bernalillo", 
  year = 2019
)
bernalillo_migration
```

## Debugging errors

To assist with debugging errors, or more generally to help users understand how **tidycensus** functions are being translated to Census API calls, **tidycensus** offers a parameter `show_call` that when set to `TRUE` prints out the actual API call that **tidycensus** is making to the Census API.

```{r}
cbsa_bachelors <- get_acs(
  geography = "cbsa",
  variables = "DP02_0068P",
  year = 2019,
  show_call = TRUE
)
```
