---
title: "04-CensusVisualization"
output: html_document
date: "2025-03-26"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Visualization

## Basics

```{r}
library(tidycensus)
```

```{r}
nm_wide <- get_acs(
  geography = "county", 
  state = "New Mexico", 
  variables = c(medinc = "B19013_001", 
                medage = "B01002_001"), 
  output = "wide",
  year = 2020
)
nm_wide
```

```{r}
library(tidyverse)
options(scipen = 999) # avoid scientific notation

ggplot(nm_wide, aes(x = medincE)) +
  geom_histogram(bins = 15)
```

```{r}
ggplot(nm_wide, aes(y = medincE)) +
  geom_boxplot()
```

## Multivariate relationships

```{r}
ggplot(nm_wide, aes(x = medageE, y = medincE)) +
  geom_point()
```

```{r}
ggplot(nm_wide, aes(x = medageE, y = medincE)) +
  geom_point() +
  geom_smooth(method = "lm")
```

## Customizing visualizations

```{r}
library(tidycensus)
library(tidyverse)
```

In this example, we will create a visualization that illustrates the percent
of commuters that take public transportation to work for the largest
metropolitan areas in the United States. The data come from the 2019 1-year
American Community Survey Data Profile, variable `DP03_0021P`.

```{r}
metros <- get_acs(
  geography = "cbsa", 
  variables = "DP03_0021P", 
  summary_var = "B01003_001", 
  survey = "acs1", 
  year = 2019
) |> 
  slice_max(summary_est, n = 20)
metros
```

```{r}
ggplot(metros, aes(x = NAME, y = estimate)) +
  geom_col()
```

> Simplify the names

```{r}
metro_plot <- metros |> 
  mutate(NAME = str_remove(NAME, "-.*$")) |> 
  mutate(NAME = str_remove(NAME, ",.*$")) |> 
  ggplot(aes(y = reorder(NAME, estimate), x = estimate)) +
  geom_col()
metro_plot
```

```{r}
metro_plot <- metro_plot +
  theme_minimal() +
  labs(title = "Public transit commute share", 
       subtitle = "2019 1-year ACS estimates",
       y = "",
       x = "ACS estimate", 
       caption = "Source: ACS Data Profile variable DP03_0021P")
metro_plot
```

```{r}
library(scales)
```

```{r}
metro_transit <- metros |> 
  mutate(NAME = str_remove(NAME, "-.*$")) %>%
  mutate(NAME = str_remove(NAME, ",.*$")) %>%
  ggplot(aes(y = reorder(NAME, estimate), x = estimate)) + 
  geom_col(color = "navy", fill = "navy", 
           alpha = 0.5, width = 0.85) +
  theme_minimal(base_size = 12, base_family = "Verdana") +
  scale_x_continuous(labels = label_percent(scale = 1)) +
  labs(title = "Public transit commute share", 
       subtitle = "2019 1-year ACS estimates",
       y = "",
       x = "ACS estimate", 
       caption = "Source: ACS Data Profile variable DP03_0021P")
metro_transit
```

## Exporting visualizations

```{r}
ggsave("metro_transit.png")
```

```{r}
ggsave(
  filename = "metro_transit.png",
  path = "./images",
  width = 8,
  height = 5, 
  units = "in", 
  dpi = 300
)
```

## Visualizing margins of error

```{r}
new_mexico <- get_decennial(
  state = "New Mexico", 
  geography = "county",
  variables = c(totalpop = "P1_001N"),
  year = 2020
) |> arrange(desc(value))
new_mexico
```

Estimates for the counties with smaller population sizes are likely to be
subject to a larger margin of error than those with larger baseline
populations. Comparing median household incomes of these sixteen counties
illustrates this point.

```{r}
nm_income <- get_acs(
  state = "New Mexico", 
  geography = "county", 
  variables = c(hhincome = "B19013_001"), 
  year = 2020
) |> 
  mutate(NAME = str_remove(NAME, " County, New Mexico"))
nm_income
```

```{r}
nm_income |> 
  ggplot(aes(x = estimate, y = reorder(NAME, estimate))) +
  geom_point(size = 2, color = "darkgreen") +
  labs(title = "Median household income", 
       subtitle = "Counties in New Mexico", 
       x = "", 
       y = "ACS estimate") +
  theme_minimal(base_size = 11) +
  scale_x_continuous(labels = label_dollar())
```

## Error bars

```{r}
nm_income |> 
  arrange(desc(moe))
```

```{r}
nm_income |> 
  ggplot(aes(x = estimate, y = reorder(NAME, estimate))) +
  geom_errorbar(aes(xmin = estimate - moe, xmax = estimate + moe)) +
  geom_point(size = 2, color = "gold") +
  theme_minimal(base_size = 10) +
  labs(title = "Median household income", 
       subtitle = "Counties in Maine", 
       x = "2016-2020 ACS estimate", 
       y = "") + 
  scale_x_continuous(labels = label_dollar())  
```

## Visualizing estimates over time

```{r}
years <- 2005:2019
names(years) <- years

bernalillo_value <- map_dfr(years, ~{
  get_acs(
    geography = "county", 
    variables = "B25077_001",
    state = "New Mexico",
    county = "Bernalillo",
    year = .x,
    survey = "acs1"
  )
}, .id = "year")
bernalillo_value
```

```{r}
bernalillo_value |> 
  ggplot(aes(x = year, y = estimate, group = 1)) +
  geom_line() +
  geom_point()
```

```{r}
santa_fe_value <- map_dfr(years, ~{
  get_acs(
    geography = "county", 
    variables = "B25077_001",
    state = "New Mexico",
    county = "Santa Fe",
    year = .x,
    survey = "acs1"
  )
}, .id = "year")
santa_fe_value |> 
  ggplot(aes(x = year, y = estimate, group = 1)) +
  geom_line() +
  geom_point()
```

```{r}
bernalillo_value |> 
  ggplot(aes(x = year, y = estimate, group = 1)) +
  geom_ribbon(aes(ymax = estimate + moe, ymin = estimate - moe),
              fill = "aquamarine", alpha = 0.4) +
  geom_line(color = "gold") +
  geom_point(color = "firebrick") +
  theme_minimal(base_size = 12) +
  scale_y_continuous(labels = label_dollar(scale = .001, suffix = "k")) +
  labs(title = "Med home value in Bernalillo County, NM", 
       x = "Year", 
       y = "ACS estimate",
       caption = "Shaded area represents ACS margin of error")
```

## Population pyramids

Population pyramids are typically constructed by visualizing population size
or proportion on the x-axis; age cohort on the y-axis; and sex is
represented categorically with male and female bars mirrored around a
central axis.

```{r}
nm <- get_estimates(
  geography = "state", 
  state = "NM", 
  product = "characteristics",
  breakdown = c("SEX", "AGEGROUP"),
  breakdown_labels = T,
  year = 2019
)
nm
```

> Processing
>
> -   remove "Both sexes" rows
> -   remove rows for "All ages"
> -   to achieve the mirroring, convert all Male values to negative

```{r}
nm_filtered <- nm |> 
  filter(str_detect(AGEGROUP, "^Age"),
         SEX != "Both sexes") |> 
  mutate(value = ifelse(SEX == "Male", -value, value))
nm_filtered
```

```{r}
nm_filtered |> 
  ggplot(aes(x = value, y = AGEGROUP, fill = SEX)) +
  geom_col()
```

-   specify appropriate chart labels
-   format the axis tick labels by using appropriate `scale_*` functions
-   set the X-axis limits to show both sides of 0 equally. In particular,
    this involves the use of custom absolute values to represent population
    sizes, and the removal of redundant age group information
-   use
    [`theme_minimal()`](https://ggplot2.tidyverse.org/reference/ggtheme.html),
    which uses a white background with muted gridlines.

```{r}
nm_pyramid <- nm_filtered |> 
  ggplot(aes(y = AGEGROUP, 
             x = value, 
             fill = SEX)) +
  geom_col(width = 0.95, alpha = 0.75) +
  theme_minimal(base_family = "Verdana", base_size = 12) +
  scale_x_continuous(
    labels = ~ number_format(scale = .001, suffix = "k")(abs(.x)),
    limits = 100000 * c(-1,1)
  ) +
  scale_y_discrete(labels = ~ str_remove_all(.x, "Age\\s|\\syears")) +
  scale_fill_manual(values = c("firebrick","aquamarine")) +
  labs(x = "", 
       y = "2019 Census Bureau population estimate", 
       title = "Population structure in New Mexico", 
       fill = "", 
       caption = "Data source: US Census Bureau population estimates & tidycensus R package")
nm_pyramid

```

## Group-wise comparisons

```{r}
housing_val <- get_acs(
  geography = "tract", 
  variables = "B25077_001",
  state = "NM", 
  county = c(
    "Bernalillo", 
    "Sandoval", 
    "Los Alamos", 
    "Santa Fe",
    "Dona Ana",
    "McKinley"
  ),
  year = 2020
)
housing_val
```

```{r}
housing_val2 <- housing_val |> 
  separate(
    NAME, into = c("tract","county","state"),
    sep = ", "
  )
housing_val2
```

```{r}
housing_val2 |> 
  group_by(county) |> 
  summarise(min = min(estimate, na.rm = T), 
            mean = mean(estimate, na.rm = T), 
            median = median(estimate, na.rm = T), 
            max = max(estimate, na.rm = T))
```

```{r}
housing_val2 |> 
  ggplot(aes(x = estimate)) +
  geom_density()
```

```{r}
housing_val2 |> 
  ggplot(aes(x = estimate, fill = county)) +
  geom_density(alpha = 0.3)
```

```{r}
housing_val2 |> 
  ggplot(aes(x = estimate)) +
  geom_density(fill = "orange", color = "red", alpha = 0.5) +
  facet_wrap(~county) +
  scale_x_continuous(label = dollar_format(scale = .001,
                                           suffix = "k")) +
  theme_minimal(base_size = 14) +
  theme(axis.text.y = element_blank(), 
        axis.text.x = element_text(angle = 45)) +
  labs(x = "ACS estimate", 
       y = "",
       title = "Median home value by census tract, 2015-2019 ACS")
```

## ggridges

```{r}
library(ggridges)
```

```{r}
housing_val2 |> 
  ggplot(aes(x = estimate, y = county)) +
  geom_density_ridges() +
  theme_ridges() +
  labs(x = "Median home value: 2016-2020 ACS estimate", 
       y = "") +
  scale_x_continuous(labels = label_dollar(scale = .001, suffix = "k", 
                                           breaks = c(0, 5e5, 1e6))) +
  theme(axis.text.x = element_text(angle = 45))
  
```

```{r}
library(ggbeeswarm)
```

```{r}
race_vars <- c(
  White = "B03002_003",
  Black = "B03002_004",
  Native = "B03002_005",
  Asian = "B03002_006",
  Hispanic = "B03002_012"
)
abq_race_income <- get_acs(
  geography = "tract", 
  state = "NM",
  county = c("Bernalillo", "Sandoval"),
  variables = race_vars,
  summary_var = "B19013_001", 
  year = 2020
)
abq_race_income <- abq_race_income |> 
  group_by(GEOID) |> 
  filter(estimate == max(estimate, na.rm = T)) |> 
  ungroup() |> 
  filter(estimate != 0)
```

```{r}
abq_race_income |> 
  ggplot(aes(x = variable, y = summary_est, color = summary_est)) +
  geom_quasirandom(alpha = 0.5) +
  coord_flip() +
  theme_minimal(base_size = 12) +
  scale_color_viridis_c(guide = "none") +
  scale_y_continuous(labels = label_dollar()) +
  labs(x = "Largest group in Census tract", 
       y = "Median household income", 
       title = "Household income distribution by largest racial/ethnic group", 
       subtitle = "Census tracts, Albuquerque Area", 
       caption = "Data source: 2016-2020 ACS")  
```

```{r}
library(geofacet)
ggplot(nm_filtered, aes(x = prop, y = AGEGROUP, fill = SEX)) + 
  geom_col(width = 1) + 
  theme_minimal() + 
  scale_fill_manual(values = c("darkred", "navy")) + 
  facet_geo(~ NAME, grid = "us_nm_counties_grid1",
            label = "code") + 
  theme(axis.text = element_blank(),
        strip.text.x = element_text(size = 8)) + 
  labs(x = "", 
       y = "", 
       title = "Population structure by age and sex", 
       fill = "", 
       caption = "Data source: US Census Bureau population estimates & tidycensus R package")
```

```{r}

```

```{r}
nm_pyramid_data <-  get_estimates(
  geography = "county",
  state = "NM",
  product = "characteristics",
  breakdown = c("SEX", "AGEGROUP"),
  breakdown_labels = TRUE,
  year = 2019
) %>%
  filter(str_detect(AGEGROUP, "^Age"),
         SEX != "Both sexes") %>%
  group_by(NAME) %>%
  mutate(prop = value / sum(value, na.rm = TRUE)) %>%
  ungroup() %>%
  mutate(prop = ifelse(SEX == "Male", -prop, prop))
nm_pyramid_data
```

```{r}
nm_pyramid_data <- nm_pyramid_data |> 
  separate(
    NAME, into = c("county","state"),
    sep = ", "
  )
nm_pyramid_data
```

```{r}
nm_pyramid_data <- nm_pyramid_data |> 
  mutate(county = str_remove(county, " County$"))
nm_pyramid_data
```

```{r}

ggplot(nm_pyramid_data, aes(x = prop, y = AGEGROUP, fill = SEX)) + 
  geom_col(width = 1) + 
  theme_minimal() + 
  scale_fill_manual(values = c("darkred", "navy")) + 
  facet_geo(~county, grid = "us_nm_counties_grid1",
            label = "county") + 
  theme(axis.text = element_blank(),
        strip.text.x = element_text(size = 8)) + 
  labs(x = "", 
       y = "", 
       title = "Population structure by age and sex", 
       fill = "", 
       caption = "Data source: US Census Bureau population estimates & tidycensus R package")
```

## Plotly

```{r}
library(plotly)
ggplotly(nm_pyramid)
```
