---
title: "05-CensusGeographicData"
output: 
  html_document:
    toc: true
date: "2025-03-26"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, paged.print = FALSE)
```

# Census Geographic Data and Applications in R

## Tigris

```{r}
library(tigris)
options(tigris_use_cache = T)
```

```{r}
states <- states()
```

```{r}
states
```

```{r}
plot(states$geometry)
```

```{r}
nm_counties <- counties("NM")
plot(nm_counties$geometry)
```

```{r}
library(sf)
```

```{r}
st_crs(nm_counties)
```

```{r}
bern_tracts <- tracts("NM", "Bernalillo")
plot(bern_tracts$geometry)
```

```{r}
sf_water <- area_water("NM", "Santa Fe")
plot(sf_water$geometry)
```

```{r}
taos_water <- area_water("NM", "Taos")
plot(taos_water$geometry)
```

```{r}
bern_water <- area_water("NM", "Bernalillo")
plot(bern_water$geometry)
```

## Points Landmarks

```{r}
nm_landmarks <- landmarks("NM", type = "point")
plot(nm_landmarks$geometry)
```

```{r}
#View(nm_landmarks)
```

## Lines

```{r}
nm_roads <- primary_secondary_roads("NM")
plot(nm_roads$geometry)
```

## Polygons

```{r}
nm_block_groups <- block_groups("NM")
plot(nm_block_groups$geometry)
```

-   `nation()`
-   [`divisions()`](https://rdrr.io/pkg/tigris/man/divisions.html)
-   [`regions()`](https://rdrr.io/pkg/tigris/man/regions.html)
-   [`states()`](https://rdrr.io/pkg/tigris/man/states.html)
-   [`counties()`](https://rdrr.io/pkg/tigris/man/counties.html)
-   [`tracts()`](https://rdrr.io/pkg/tigris/man/tracts.html)
-   [`block_groups()`](https://rdrr.io/pkg/tigris/man/block_groups.html)
-   [`blocks()`](https://rdrr.io/pkg/tigris/man/blocks.html)
-   [`places()`](https://rdrr.io/pkg/tigris/man/places.html)
-   [`pumas()`](https://rdrr.io/pkg/tigris/man/pumas.html)
-   [`school_districts()`](https://rdrr.io/pkg/tigris/man/school_districts.html)
-   [`zctas()`](https://rdrr.io/pkg/tigris/man/zctas.html)
-   [`congressional_districts()`](https://rdrr.io/pkg/tigris/man/congressional_districts.html)
-   [`state_legislative_districts()`](https://rdrr.io/pkg/tigris/man/state_legislative_districts.html)
-   [`voting_districts()`](https://rdrr.io/pkg/tigris/man/voting_districts.html)
-   [`area_water()`](https://rdrr.io/pkg/tigris/man/area_water.html)
-   [`linear_water()`](https://rdrr.io/pkg/tigris/man/linear_water.html)
-   [`coastline()`](https://rdrr.io/pkg/tigris/man/coastline.html)
-   [`core_based_statistical_areas()`](https://rdrr.io/pkg/tigris/man/core_based_statistical_areas.html)
-   [`combined_statistical_areas()`](https://rdrr.io/pkg/tigris/man/combined_statistical_areas.html)
-   [`metro_divisions()`](https://rdrr.io/pkg/tigris/man/metro_divisions.html)
-   [`new_england()`](https://rdrr.io/pkg/tigris/man/new_england.html)
-   [`county_subdivisions()`](https://rdrr.io/pkg/tigris/man/county_subdivisions.html)
-   [`urban_areas()`](https://rdrr.io/pkg/tigris/man/urban_areas.html)
-   [`primary_roads()`](https://rdrr.io/pkg/tigris/man/primary_roads.html)
-   [`primary_secondary_roads()`](https://rdrr.io/pkg/tigris/man/primary_secondary_roads.html)
-   [`roads()`](https://rdrr.io/pkg/tigris/man/roads.html)
-   [`rails()`](https://rdrr.io/pkg/tigris/man/rails.html)
-   [`native_areas()`](https://rdrr.io/pkg/tigris/man/native_areas.html)
-   [`alaska_native_regional_corporations()`](https://rdrr.io/pkg/tigris/man/alaska_native_regional_corporations.html)
-   [`tribal_block_groups()`](https://rdrr.io/pkg/tigris/man/tribal_block_groups.html)
-   [`tribal_census_tracts()`](https://rdrr.io/pkg/tigris/man/tribal_census_tracts.html)
-   [`tribal_subdivisions_national()`](https://rdrr.io/pkg/tigris/man/tribal_subdivisions_national.html)
-   [`landmarks()`](https://rdrr.io/pkg/tigris/man/landmarks.html)
-   [`military()`](https://rdrr.io/pkg/tigris/man/military.html)

```{r}
library(ggplot2)
ggplot(bern_tracts) +
  geom_sf()
```

```{r}
ggplot(bern_tracts) +
  geom_sf() +
  theme_void()
```

```{r}
library(patchwork)

sf_block_groups <- block_groups("NM", "Santa Fe")
sf_tracts <- tracts("NM", "Santa Fe")
```

```{r}
gg1 <- ggplot(sf_tracts) +
  geom_sf() +
  theme_void() +
  labs(title = "Census tracts")

gg2 <- ggplot(sf_block_groups) +
  geom_sf() +
  theme_void() +
  labs(title = "Block groups")

gg1 + gg2
```

```{r}
library(mapview)
mapview(sf_tracts)
```

## Cartographic boundary shapefiles

Most polygon datasets in **tigris** are available as cartographic boundary
files, accessible with the argument `cb = TRUE`. These are generalized in
the interior and clipped to the shoreline. When using the `cb = TRUE`
argument with counties and larger geographies, users can also specify one of
three resolutions with the `resolution` argument: `"500k"` (the default),
`"5m"`, or `"20m"`, with higher values representing more generalized
boundaries and smaller file sizes.

```{r}
mi_counties <- counties("MI")
mi_counties_cb <- counties("MI", cb = TRUE)

mi_tiger_gg <- ggplot(mi_counties) + 
  geom_sf() + 
  theme_void() + 
  labs(title = "TIGER/Line")

mi_cb_gg <- ggplot(mi_counties_cb) + 
  geom_sf() + 
  theme_void() + 
  labs(title = "Cartographic boundary")

mi_tiger_gg + mi_cb_gg
```

## Cache directory

```{r}
options(tigris_use_cache = T)

rappdirs::user_cache_dir("tigris")
```

## Yearly difference in TIGER/Line files

```{r}
library(glue)
library(purrr)
yearly_plots <- map(seq(1990, 2020, 10), ~{
  year_tracts <- tracts("TX", "Tarrant", year = .x, 
                        cb = T)
  
  ggplot(year_tracts) +
    geom_sf() +
    theme_void() +
    labs(title = glue("{.x}: {nrow(year_tracts)} tracts"))
})
```

```{r}
(yearly_plots[[1]] + yearly_plots[[2]]) /
  (yearly_plots[[3]] + yearly_plots[[4]])
```

```{r}
yearly_plots <- map(seq(1990, 2020, 10), ~{
  year_tracts <- tracts("NM", "Bernalillo", year = .x, 
                        cb = T)
  
  ggplot(year_tracts) +
    geom_sf() +
    theme_void() +
    labs(title = glue("{.x}: {nrow(year_tracts)} tracts"))
})
(yearly_plots[[1]] + yearly_plots[[2]]) /
  (yearly_plots[[3]] + yearly_plots[[4]])
```

## Combining tigris datasets

For years starting 2019, `cb = TRUE` obtains all blockgroups.

```{r}
us_bgs_2020 <- block_groups(cb = T, year = 2020)
nrow(us_bgs_2020)
```

> for years prior to 2019

```{r}
state_codes <- c(state.abb, "DC", "PR")

us_bgs_2018 <- map_dfr(
  state_codes, 
  ~block_groups(
    state = .x, 
    cb = T, 
    year = 2018
  )
)

nrow(us_bgs_2018)
```

## CRS

```{r}
library(sf)

st_crs(nm_counties)
```

## crsuggest

```{r}
library(crsuggest)

nm_crs <- suggest_crs(nm_counties)
nm_crs
```

```{r}
nm_projected <- st_transform(nm_counties, crs = 32113)
head(nm_projected)
```

## `coord_sf()`

```{r}
library(ggplot2)
library(patchwork)
options(scipen = 999)

p1 <- ggplot(nm_counties) +
  geom_sf() +
  coord_sf(crs = 32113)
p2 <- ggplot(nm_counties) +
  geom_sf() +
  coord_sf(crs = 32113, datum = 32113)

p1 + p2
```

## Shifting and rescaling for national US mapping

```{r}
us_states <- states(cb = T, resolution = "20m")

ggplot(us_states) +
  geom_sf() +
  theme_void()
```

```{r}
ggplot(us_states) +
  geom_sf() +
  coord_sf(crs = "ESRI:102003") +
  theme_void()
```

> with default `preserve_area = TRUE` and `position = "below"`

```{r}
us_states_shifted <- shift_geometry(us_states)

ggplot(us_states_shifted) +
  geom_sf() +
  theme_void()
```

```{r}
us_states_outside <- shift_geometry(us_states, 
                                    preserve_area = T, 
                                    position = "outside")

ggplot(us_states_outside) +
  geom_sf() +
  theme_void()
```

## Polygons to points

```{r}
library(tidyverse)
nm_places <- places("NM", cb = T)|> 
  filter(NAME %in% c("Albuquerque", "Rio Rancho", "Farmington", 
                     "Gallup", "Santa Fe", "Las Cruces", 
                     "Roswell", "Santa Rosa", "Raton")) |> 
  st_transform(crs = 32113)

nm_outline <- states(cb = T) |> 
  filter(NAME == "New Mexico") |> 
  st_transform(crs = 32113)

ggplot() +
  geom_sf(data = nm_outline) +
  geom_sf(data = nm_places, fill = "red", color = NA) +
  theme_void()
```

```{r}
nm_centroids <- st_centroid(nm_places)

ggplot() +
  geom_sf(data = nm_outline) +
  geom_sf(data = nm_centroids, color = "red", xize = 3) +
  theme_void()
```

## Exploding Multipolygons

```{r}
fl_counties <- counties("FL", cb = TRUE)
fl_projected <- st_transform(fl_counties, crs = 3087)
lee <- fl_projected |> 
  filter(NAME == "Lee")
```

```{r}
mapview(lee)
```

```{r}
lee
```

```{r}
lee_seperate <- st_cast(lee, "POLYGON")
lee_seperate
```

```{r}
lee_seperate[2,] |> 
  mapview()
```
