---
title: "06.2-CartographicWorkflows"
output: 
  html_document:
    toc: true
date: "2025-03-28"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, paged.print = FALSE)
```

# Cartographic workflows with non-Census Data

## National election mapping with tigris shapes

```{r}
library(tidyverse)
library(tigris)
options(tigris_use_cache = TRUE)
```

```{r}
# Data source: https://cookpolitical.com/2020-national-popular-vote-tracker
vote2020 <- read_csv("data/votes.csv")
```

```{r}
names(vote2020)
```

```{r}
spec(vote2020)
```

```{r}
us_states <- states(cb = T, resolution = "20m") |> 
  filter(NAME != "Puerto Rico") |> 
  shift_geometry()

us_states_joined <- us_states |> 
  left_join(vote2020, by = c("NAME" = "state"))
```

Before proceeding we’ll want to do some quality checks. In
[`left_join()`](https://dplyr.tidyverse.org/reference/mutate-joins.html),
values must match exactly between `NAME` and `state` to merge correctly -
and this is not always guaranteed when using data from different sources.
Let’s check to see if we have any problems:

```{r}
table(is.na(us_states_joined$state))
```

```{r}
setdiff(vote2020$state, us_states$NAME)
```

```{r}
us_states_joined |> ggplot(aes(fill = called)) +
  geom_sf(color = "white", lwd = 0.2) +
  scale_fill_manual(values = c("blue", "red")) +
  theme_void() +
  labs(fill = "Party", 
       title = "2020 Presidential Election Results", 
       caption = "Note: Nebraska and Maine split electoral college votes by congressional district")
```

## Working with ZCTAs (zip codes)

The US Census Bureau allows for an approximation of zip code mapping with
Zip Code Tabulation Areas, or ZCTAs. ZCTAs are shapes built from Census
blocks in which the most common zip code for addresses in each block
determines how blocks are allocated to corresponding ZCTAs. While ZCTAs are
not recommended for spatial analysis due to these irregularities, they can
be useful for visualizing data distributions when no other granular
geographies are available.

An example of this is the [Internal Revenue Service’s Statistics of Income
(SOI)
data](https://www.irs.gov/statistics/soi-tax-stats-individual-income-tax-statistics-2018-zip-code-data-soi),
which includes a wide range of indicators derived from tax returns. The most
detailed geography available is the zip code level in this dataset, meaning
that within-county visualizations require using ZCTAs. Let’s read in the
data for 2018 from the IRS website:

```{r}
irs_data <- read_csv("https://www.irs.gov/pub/irs-soi/18zpallnoagi.csv")
ncol(irs_data)
```

The dataset contains 153 columns [which are identified in the linked
codebook](https://www.irs.gov/pub/irs-soi/18zpdoc.docx). Geographies are
identified by the `ZIPCODE` column, which shows aggregated data by state
(`ZIPCODE == "000000"`) and by zip code. We might be interested in
understanding the geography of self-employment income within a given region.
We’ll retain the variables `N09400`, which represents the number of tax
returns with self-employment tax, and `N1`, which represents the total
number of returns.

```{r}
self_employment <- irs_data |> 
  select(ZIPCODE, self_emp = N09400, total = N1)
```

```{r}
library(mapview)
library(tigris)
options(tigris_use_cache = T)

abq_zctas <- zctas(
  cb = T,
  starts_with = c("871"),
  year = 2018
)

mapview(abq_zctas)
```

```{r}
names(abq_zctas)
```

```{r}
abq_se_data <- abq_zctas |> 
  left_join(self_employment, by = c("GEOID10" = "ZIPCODE")) |> 
  mutate(pct_se = 100 * self_emp / total) |> 
  select(GEOID10, self_emp, pct_se)
abq_se_data
```

```{r}
library(tmap)
tm_shape(abq_se_data, projection = 32113) +
  tm_fill(col = "pct_se", palette = "Purples") +
  tm_title("% self-employed\n2010 IRS Data")
```

The choropleth map shows that self-employment filings are more common in
suburban Boston ZCTAs than nearer to the urban core, generally speaking.
However, we might also be interested in understanding where most
self-employment income filings are located rather than their share relative
to the total number of returns filed. This requires visualizing the
`self_emp` column directly. As discussed earlier in this chapter, a
graduated symbol map with
[`tm_bubbles()`](https://rdrr.io/pkg/tmap/man/tm_symbols.html) is preferable
to a choropleth map for this purpose.

```{r}
tm_shape(abq_se_data) +
  tm_polygons() +
  tm_bubbles(size = "self_emp", fill_alpha = 0.5, 
             col = "navy", 
             title.size = "Self-employed filers,\n2018 IRS SOI data")
```

# Interactive mapping

Let’s start by getting some illustrative data on the percentage of the
population aged 25 and up with a bachelor’s degree or higher from the
2016-2020 ACS.

```{r}
library(tidycensus)
bern_bachelors <- get_acs(
  geography = "tract",
  variables = "DP02_0068P",
  state = "NM", 
  county = "Bernalillo",
  year = 2020,
  geometry = T
)
```

```{r}
library(mapview)
mapview(bern_bachelors, zcol = "estimate")
```

```{r}
ttm()
tm_shape(bern_bachelors) +
  tm_fill(col = "estimate", palette = "magma",
          alpha = 0.5)
ttm()
```

## Leaflet

```{r}
library(leaflet)

pal <- colorNumeric(
  palette = "magma",
  domain = bern_bachelors$estimate
)

pal(c(10, 20, 30, 40, 50))
```

```{r}
leaflet() |> 
  addProviderTiles(providers$Stadia.StamenTonerLite) |> 
  addPolygons(data = bern_bachelors, 
              color = ~pal(estimate),
              weight = 0.5, 
              smoothFactor = 0.2,
              fillOpacity = 0.5,
              label = ~estimate) |> 
  addLegend(
    position = "bottomright",
    pal = pal,
    values = bern_bachelors$estimate,
    title = "% with bachelor's<br/>degree"
  )
```

```{r}
library(ggiraph)
library(scales)

us_value <- get_acs(
  geography = "state",
  variables = "B25077_001",
  year = 2019,
  survey = "acs1",
  geometry = TRUE,
  resolution = "20m"
)

us_value_shifted <- us_value |> 
  shift_geometry(position = "outside") |> 
  mutate(tooltip = paste(NAME, estimate, sep = ": "))

gg <- ggplot(us_value_shifted, aes(fill = estimate)) +
  geom_sf_interactive(aes(tooltip = tooltip, data_id = NAME), 
                      size = 0.1) +
  scale_fill_viridis_c(option = "plasma", labels = label_dollar()) +
  labs(title = "Median housing value by State, 2019",
       caption = "Data source: 2019 1-year ACS, US Census Bureau",
       fill = "ACS estimate") + 
  theme_void()   

gg
```

```{r}
girafe(ggobj = gg) |> 
  girafe_options(opts_hover(css = "fill:cyan"),
                 opts_zoom(max = 10))
```

## Migration flows

```{r}
abq_inflows <- get_flows(
  geography = "county",
  state = "NM",
  county = "Bernalillo",
  geometry = T
) |> 
  filter(variable == "MOVEDIN") |> 
  na.omit() |> 
  arrange(desc(estimate))
abq_inflows
```

```{r}
library(mapdeck)

key <- "pk.eyJ1IjoiYmlzY290dHkiLCJhIjoiY2x0azc2eGo5MHhkYTJqb3lndG44bjN3OSJ9.KVJqBYZMqDW_E9JisfCiLg"

abq_inflows |> 
  slice_max(estimate, n = 30) |> 
  mutate(weight = estimate / 500) |> 
  mapdeck(token = key) |> 
  add_arc(origin = "centroid2",
          destination = "centroid1",
          stroke_width = "weight", 
          update_view = F)
```

## Linking maps and charts

### Choropleth with margin of error plot

```{r}
library(tidycensus)
library(ggiraph)
library(tidyverse)
library(patchwork)
library(scales)
```

```{r}
nm_income <- get_acs(
  geography = "county",
  variables = "B19013_001",
  state = "NM",
  year = 2020,
  geometry = T
) |> 
  mutate(NAME = str_remove(NAME, " County, New Mexico"))
```

```{r}
nm_map <- ggplot(nm_income, aes(fill = estimate)) +
  geom_sf_interactive(aes(data_id = GEOID)) +
  scale_fill_distiller(palette = "Reds",
                       direction = 1,
                       guide = "none") +
  theme_void()
```

```{r}
nm_plot <- ggplot(nm_income, aes(x = estimate,
                                 y = reorder(NAME, estimate),
                                 fill = estimate)) +
  geom_errorbar(aes(xmin = estimate - moe, xmax = estimate + moe)) +
  geom_point_interactive(aes(data_id = GEOID),
                         color = "black", size = 4, shape = 21) +
  scale_fill_distiller(palette = "Reds", direction = 1,
                       labels = label_dollar()) +
  scale_x_continuous(labels = label_dollar()) +
  labs(title = "New Mexico household income by county",
       subtitle = "2016-2020 American Community Survey",
       y = "", x = "ACS estimate (bars represent margin of error",
       fill = "ACS estimate") +
  theme_minimal(base_size = 14)
```

```{r}
library(patchwork)
nm_map + nm_plot
```

```{r}
girafe(ggobj = nm_map + nm_plot,
       width_svg = 10, height_svg = 5) |> 
  girafe_options(opts_hover(css = "fill:cyan;"))
```
