---
title: "08.2-Regression Modelling Census Data"
output: 
  md_document:
    variant: gfm
date: "2025-03-28"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, paged.print = FALSE)
```

# 08.2-Regression Modeling Census Data

Demographic statistics are typically highly correlated, and spatial
information tends to be autocorrelated spatially, so many predictors are not
independent.

This exercise looks at median home values.

## Data setup and esda

```{r}
library(tidycensus)
library(sf)
library(tidyverse)

nm_counties <- c("Valencia", "Bernalillo","Sandoval", 
                 "Santa Fe", "Los Alamos")

variables_to_get <- c(
  median_value = "B25077_001",
  median_rooms = "B25018_001",
  median_income = "DP03_0062",
  total_population = "B01003_001",
  median_age = "B01002_001",
  pct_college = "DP02_0068P",
  pct_foreign_born = "DP02_0094P",
  pct_white = "DP05_0077P",
  median_year_built = "B25037_001",
  percent_ooh = "DP04_0046P"
)

abq_data <- get_acs(
  geography = "tract", 
  variables = variables_to_get,
  state = "NM",
  geometry = T,
  county = nm_counties,
  output = "wide",
  year = 2020
) |> 
  select(-NAME) |> 
  st_transform(32113)

head(abq_data)
```

## Visualization

```{r}
library(tidyverse)
library(patchwork)

mhv_map <- ggplot(abq_data, aes(fill = median_valueE)) +
  geom_sf(color = NA) +
  scale_fill_viridis_c(labels = scales::label_dollar()) +
  theme_void() +
  labs(fill = "Median home value")

mhv_histogram <- ggplot(abq_data, aes(x = median_valueE)) +
  geom_histogram(alpha = 0.5, fill = "navy", color = "navy",
                 bins = 40) +
  theme_minimal() +
  scale_x_continuous(labels = scales::label_number(accuracy = 0.1)) +
  labs(x = "Median home value")

mhv_map + mhv_histogram
```

> to make the right-skewed distribution more normal, apply log
> transformation

```{r}
mhv_map_log <- ggplot(abq_data, aes(fill = log(median_valueE))) +
  geom_sf(color = NA) +
  scale_fill_viridis_c() +
  theme_void() +
  labs(fill = "Median home\nvalue (log)")

mhv_histogram_log <- ggplot(abq_data, aes(x = log(median_valueE))) +
  geom_histogram(alpha = 0.5, fill = "navy", color = "navy",
                 bins = 40) +
  theme_minimal() +
  scale_x_continuous(labels = scales::label_number(accuracy = 0.1)) +
  labs(x = "Median home value (log)")

mhv_map_log + mhv_histogram_log
```

## Feature engineering

> add population density and median age of housing structures

```{r}
library(units)

abq_data_for_model <- abq_data |> 
  mutate(pop_density = as.numeric(set_units(total_populationE / st_area(geometry), "1/km2")),
         median_structure_age = 2018 - median_year_builtE)  |> 
    select(!ends_with("M")) |> 
  rename_with(.fn = ~str_remove(.x, "E$")) |> 
  na.omit()
```

```{r}
head(abq_data_for_model, 3)
```

## First regression model

The formula can be supplied as a string or unquoted in the function call.

```{r}
formula <- "log(median_value) ~ median_rooms + median_income + pct_college + pct_foreign_born + pct_white + median_age + median_structure_age + percent_ooh + pop_density + total_population"

model1 <- lm(formula = formula, data = abq_data_for_model)

summary(model1)
```

> percent going to college and median age are positively correlated, while
> percentage foreign born and non-hispanic white is negatively correlated.
> R2 value suggests that the model accounts for \~73% of the variance of
> median_value.
>
> Check for collinearity

```{r}
library(corrr)

abq_estimates <- abq_data_for_model |> 
  select(-GEOID, -median_value, -median_year_built) |> 
  st_drop_geometry()

correlations <- correlate(abq_estimates, method = "pearson")
```

```{r}
network_plot(correlations)
```

Collinearity can be diagnosed further by calculating the *variance inflation
factor*
(VIF) for the model, which takes into account not just pairwise
correlations but the extent to which predictors are collinear with all
other predictors. A VIF value of 1 indicates no collinearity; VIF values
above 5 suggest a level of collinearity that has a problematic
influence on model interpretation.

```{r}
library(car)

vif(model1)
```

None are above 5, otherwise can remove and re-run. I'll remove the highest.

```{r}
formula2 <- "log(median_value) ~ median_income + pct_college + pct_foreign_born + pct_white + median_age + median_structure_age + percent_ooh + pop_density + total_population"

model2 <- lm(formula = formula2, data = abq_data_for_model)

summary(model2)
```

> R value doesn't change but median_income becomes a stronger predictor.

```{r}
vif(model2)
```

## Dimension reduction (PCA)

```{r}
pca <- prcomp(
  formula = ~.,
  data = abq_estimates,
  scale. = TRUE,
  center = TRUE
)

summary(pca)
```

> relate components back to variables

```{r}
pca_tibble <- pca$rotation |> 
  as_tibble(rownames = "predictor")
pca_tibble
```

```{r}
pca_tibble |> 
  select(predictor:PC5) |> 
  pivot_longer(PC1:PC5, names_to = "component", values_to = "value") |> 
  ggplot(aes(x = value, y = predictor)) +
  geom_col(fill = "darkgreen", color = "darkgreen", alpha = 0.5) +
  facet_wrap(~component, nrow = 1) +
  labs(y = NULL, x = "Value") +
  theme_minimal()
```

`percent_ooh`, `pct_white`, `pct_college`, `median_rooms`, `median_income`
and `median_age` load negatively, `pop_density` and `foregn_born` load
positively.

```{r}
components <- predict(pca, abq_estimates)

abq_pca <- abq_data_for_model |> 
  select(GEOID, median_value) |> 
  cbind(components)

ggplot(abq_pca, aes(fill = PC1)) +
  geom_sf(color = NA) +
  theme_void() +
  scale_fill_viridis_c()
```

```{r}
pca_formula <- paste0("log(median_value) ~ ",
                      paste0('PC', 1:6, collapse = ' + '))

pca_model <- lm(formula = pca_formula, data = abq_pca)

summary(pca_model)
```
