---
title: "08.3-SpatialRegression"
output: 
  md_document:
    variant: gfm
date: "2025-04-19"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, paged.print = FALSE)
```

```{r}
library(tidycensus)
library(tidyverse)
library(sf)
```

# Spatial regression

> check residuals for model

```{r}
nm_counties <- c("Valencia", "Bernalillo","Sandoval", 
                 "Santa Fe", "Los Alamos")

variables_to_get <- c(
  median_value = "B25077_001",
  median_rooms = "B25018_001",
  median_income = "DP03_0062",
  total_population = "B01003_001",
  median_age = "B01002_001",
  pct_college = "DP02_0068P",
  pct_foreign_born = "DP02_0094P",
  pct_white = "DP05_0077P",
  median_year_built = "B25037_001",
  percent_ooh = "DP04_0046P"
)

abq_data <- get_acs(
  geography = "tract", 
  variables = variables_to_get,
  state = "NM",
  geometry = T,
  county = nm_counties,
  output = "wide",
  year = 2020
) |> 
  select(-NAME) |> 
  st_transform(32113)

library(units)

abq_data_for_model <- abq_data |> 
  mutate(pop_density = as.numeric(set_units(total_populationE / st_area(geometry), "1/km2")),
         median_structure_age = 2018 - median_year_builtE)  |> 
    select(!ends_with("M")) |> 
  rename_with(.fn = ~str_remove(.x, "E$")) |> 
  na.omit()

formula1 <- "log(median_value) ~ median_rooms + median_income + pct_college + pct_foreign_born + pct_white + median_age + median_structure_age + percent_ooh + pop_density + total_population"

model1 <- lm(formula = formula1, data = abq_data_for_model)

summary(model1)
```

```{r}
abq_data_for_model$residuals <- residuals(model1)

ggplot(abq_data_for_model, aes(x = residuals)) +
  geom_histogram(bins = 50, alpha = 0.5, 
                 color = "navy", fill = "navy") +
  theme_minimal()
```

> Evaluate the degree that the model's performance depends on spatial
> autocorrelation

```{r}
library(spdep)

wts <- abq_data_for_model |> 
  poly2nb() |> 
  nb2listw()

moran.test(abq_data_for_model$residuals, wts)
```

> Low but statistically significant

```{r}
abq_data_for_model$lagged_residuals <- lag.listw(wts, abq_data_for_model$residuals)

ggplot(abq_data_for_model, aes(x = residuals, y = lagged_residuals)) +
  theme_minimal() +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", color = "red")
```

> The residuals have positive autocorrelation, so must use spatial
> regression methods

## Spatial regression methods

> Spatial lag models account for \*spatial spillover effects\*, spatial
> error models include a spatial lag in the error term to account for
> spatial processes in the residuals.

### Spatial lag models

By including a spatial lag of the outcome variable, the model accounts for
spatial spillover effects.

$$
Y_i=\alpha+\rho Y_{lag−i}+\sum_k \beta_kX_{ki}+ϵ_i
$$ where

$$
Y_{lag−i}=\sum_j w_{ij}Y_j
$$

```{r}
library(spatialreg)

lag_model <- lagsarlm(
  formula = formula1,
  data = abq_data_for_model,
  listw = wts
)

summary(lag_model, Nagelkerke = TRUE)
```

Similar results but much less dramatic. Rho is positive and statistically
significant, indicative of spatial spillover effects.

### Spatial error models

$$
Y_i=\alpha + \sum_k \beta_kX_{ki}+u_i
$$

where

$$
u_i=\lambda u_{lag−i}+\epsilon
$$

and

$$
u_{lag−i}=\sum_j w_{ij}u_j
$$

```{r}
error_model <- errorsarlm(
  formula = formula1,
  data = abq_data_for_model,
  listw = wts
)

summary(error_model, Nagelkirke = T)
```

> large and significant $\lambda$ illustrates the importance of accounting
> for spatial autocorrelation.

## Selecting the model

The two types of models can also be evaluated with respect to some
quantitative metrics. For example, we can re-compute Moran’s I over the
model residuals to see if the spatial model has resolved our problems with
spatial dependence

```{r}
moran.test(lag_model$residuals, wts)
```

```{r}
moran.test(error_model$residuals, wts)
```

> Both models reduce Moran's $I$, but the error model does better.

```{r}
lm.RStests(
  model1,
  wts,
  test = c("LMerr", "LMlag", "RLMerr", "RLMlag")
)
```

All are large an statistically significant.
