---
title: "09-CensusMicrodata"
output: 
  md_document:
    variant: gfm
date: "2025-04-19"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, paged.print = FALSE)
```

# Census Microdata

## Basic usage

```{r}
library(tidycensus)
```

Request 1-year 2023 ACS with sex, age and household type

```{r}
nm_pums <- get_pums(
  variables = c("SEX", "AGEP", "HHT"),
  state = "NM",
  survey = "acs1",
  year = 2023
)
head(nm_pums)
```

```{r}
nrow(nm_pums)
```

-   `SERIALNO`: a serial number that uniquely identifies households in the
    sample;
-   `SPORDER`: the order of the person in the household, which when combined
    with `SERIALNO` uniquely identifies a person;
-   `WGTP`: the household weight;
-   `PWGTP`: the person weight;
-   `ST`: the state FIPS code.

In general, the weights can be interpreted as the number of observations
represented by the row, so `PWGPT` is the number of people.

Inferences about population characteristics can be made by summing over
the weights columns. For example, let’s say we want to get an estimate
of the number of people in New Mexico who are 50 years old in 2019, and
compare this with the total population in New Mexico.

```{r}
library(tidyverse)

nm_age_50 <- filter(nm_pums, AGEP == 50)

print(sum(nm_pums$PWGTP))
```

```{r}
print(sum(nm_age_50$PWGTP))
```

> There is a column for households and one for persons

```{r}
nm_hh_example <- filter(nm_pums, SERIALNO == "2023HU0000811")
nm_hh_example
```

Household includes two males ages 37 and 16, three females 37, 9 and 21. HHT
of 1 means married couple household.

To determine the number of households in the dataset, sum the WGTP column
per household.

```{r}
nm_households <- filter(nm_pums, SPORDER == 1)

sum(nm_households$WGTP)
```

### Including vacant units

```{r}
nm_with_vacant <- get_pums(
  variables = c("SEX", "AGEP", "HHT"),
  state = "NM",
  survey = "acs1",
  year = 2023,
  return_vacant = TRUE
) |> 
  arrange(VACS)
```

```{r}
head(nm_with_vacant)
```

Vacant housing units do not have person-level characteristics

## Working with PUMS variables

```{r eval=FALSE}
View(pums_variables)
```

-   `var_code` gives you the variable codes that should be supplied to the
    `variables` parameter (as a character vector) in
    [`get_pums()`](https://walker-data.com/tidycensus/reference/get_pums.html).
    These variables will be represented in the columns of your output
    dataset.
-   `var_label` is a more informative description of the variable’s topic.
-   `data_type` is one of `"chr"`, for categorical variables that will be
    returned as R character strings, or `"num"`, for variables that will be
    returned as numeric.
-   `val_min` and `val_max` provide information about
    the meaning of the data values. For categorical variables, these two
    columns will be the same; for numeric variables, they will give you the
    possible range of data values.
-   `val_label` contains the value labels, which are particularly important
    for understanding the content of categorical variables.

### Recoding PUMS variables

Get additional contextual information

```{r}
nm_pums_recoded <- get_pums(
  variables = c("SEX", "AGEP", "HHT"),
  state = "NM",
  survey = "acs1",
  year = 2023,
  recode = TRUE
)
head(nm_pums_recoded)
```

### Variable filters

To reduce the size of the data returned, can subset by filtering variables.
For example to return only women between 30 and 49 from 5-year PUMS

```{r}
nm_pums_filtered <- get_pums(
  variables = c("SEX", "AGEP", "HHT"),
  state = "NM",
  survey = "acs5",
  variables_filter = list(
    SEX = 2,
    AGEP = 30:49
  ),
  year = 2023
)
head(nm_pums_filtered)
```

```{r}
wy_pums_filtered <- get_pums(
  variables = c("SEX", "AGEP", "HHT"),
  state = "WY",
  survey = "acs5",
  variables_filter = list(
    SEX = 2,
    AGEP = 30:49
  ),
  year = 2019
)
```

## Public Use Microdata Areas

Smallest available geographies identiifiable in PUMS datasets, usually
between 100k and 200k people, redrawn with each census.

```{r}
library(tigris)
options(tigris_use_cache = T)

nm_pumas <- pumas(state = "NM", cb = T, year = 2020)

nm_pumas |> 
  ggplot() +
  geom_sf() +
  theme_void()
```

```{r}
nrow(nm_pumas)
```

```{r}
unique(nm_pumas$NAMELSAD20)
```

## Working with PUMAs in PUMS data

```{r}
nm_age_by_puma <- get_pums(
  variables = c("PUMA", "AGEP"),
  state = "NM",
  survey = "acs5",
  year = 2023
)
head(nm_age_by_puma)
```

PUMA IDs are replicated across states, so the `PUMA` column should be
combined with the `ST` column to uniquely identify PUMAs when performing
multi-state analyses.

The `puma` argument in
[`get_pums()`](https://walker-data.com/tidycensus/reference/get_pums.html)
can also be used to obtain data for a specific PUMA or multiple PUMAs. Like
the `variables_filter` parameter, `puma` uses a query on the API side to
reduce long download times for users only interested in a geographical
subset of data.

```{r}
nm_puma_subset <- get_pums(
  variables = "AGEP",
  state = "NM",
  survey = "acs5",
  puma = "00400",
  year = 2023
)
nm_puma_subset
```

### Multi-state queries

```{r}
two_state_puma_subset <- get_pums(
  variables = "AGEP",
  state = "multiple",
  survey = "acs5",
  puma = c("NM" = "00400", "AZ" = "00400"),
  year = 2023
)
two_state_puma_subset
```
