---
title: "11.1-NHGIS"
output: 
  md_document:
    variant: gfm
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, paged.print = FALSE)
```

# Working with NHGIS historical data

The example in this section will illustrate an applied workflow using NHGIS and its companion R package, **ipumsr** ([Ellis and Burk 2020](https://walker-data.com/census-r/references.html#ref-ellis_and_burk2020)) to map geographies of immigration in New York City from the 1910 Census.

```{r message=FALSE}
library(ipumsr)
library(tidyverse)
```

```{r}
nyc_1910_shp <- read_ipums_sf(
  "data/nhgis0001_shape/nhgis0001_shapefile_tl2000_us_tract_1910.zip",
  file_select = "US_tract_1910.shp")

nyc_data <- read_nhgis("data/nhgis0001_csv.zip")
```

```{r}
nyc_1910 <- ipums_shape_left_join(
  nyc_data,
  nyc_1910_shp,
  by = "GISJOIN"
) |> 
  filter(str_detect(GISJOIN, "G36"))
```

```{r}
nyc_1910
```

```{r eval=FALSE}
View(sf::st_drop_geometry(nyc_1910))
```

```{r}
library(sf)
nyc_1910_p <- st_transform(nyc_1910, 6538)
plot(nyc_1910_p$geometry)
```

As there is no “total population” column in the dataset, the code below uses **dplyr**’s [`rowwise()`](https://dplyr.tidyverse.org/reference/rowwise.html) and [`c_across()`](https://dplyr.tidyverse.org/reference/c_across.html) functions to perform row-wise calculations and sum across the columns `A60001` through `A60007`.

Interestingly, there were only three races, white, black and other. A60005 is described "foreign-born white". There is no corresponding category for non-white.

```{r}
nyc_pctfb <- nyc_1910_p |> 
  rowwise() |> 
  mutate(total = sum(c_across(A60001:A60007))) |> 
  ungroup() |> 
  transmute(
    tract_id = GISJOIN,
    pct_fb = A60005 / total
  )
```

```{r}
ggplot(nyc_pctfb, aes(fill = pct_fb)) +
  geom_sf(color = NA) +
  scale_fill_viridis_c(option = "rocket",
                       labels = scales::percent,
                       direction = -1) +
  theme_void(base_family = "Verdana") + 
  labs(title = "Percent foreign-born by Census tract, 1910",
       subtitle = "New York City",
       caption =  "Data source: NHGIS",
       fill = "Percentage")  
```

For example, the number of Italy-born residents by Census tract is represented in the column `A6G014`

```{r}
library(tidycensus)

italy_dots <- nyc_1910_p |> 
  as_dot_density(
    value = "A6G014",
    values_per_dot = 100
  )

nyc_base <- nyc_1910_p |> 
  st_union()
```

```{r}
ggplot() +
  geom_sf(data = nyc_base, size = 0.1) +
  geom_sf(data = italy_dots, shape = ".", color = "darkgreen") +
  theme_void(base_family = "Verdana") +
  labs(title = "Italy-born population in New York City, 1910",
       subtitle = "1 dot = 100 people",
       caption = "Data source: NHGIS")
```

```{r}
nyc_data$GISJOIN
```
