---
title: "11.2-CompleteCountMicrodata"
output: 
  md_document:
    variant: gfm
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, paged.print = FALSE)
```

# Complete Count Historical Microdata

```{r}
library(RPostgres)
library(tidyverse)
library(dbplyr)
```

## Loading data

```{r}
conn <- dbConnect(
  drv = Postgres(),
  user = "census",
  password = "",
  host = "localhost",
  port = "5432",
  dbname = "census"
)
```

```{r}
census1910 <- tbl(conn, in_schema("ipums", "usa"))

census1910
```

## Analyzing big Census microdata

```{r}
census1910 |> summarize(n())
```

```{r}
census1910 |> 
  filter(AGE > 17, STATEFIP == "48") |> 
  group_by(SEX, LIT) |> 
  summarise(num = n())
```

```{r}
census1910 |> 
  filter(AGE > 17, STATEFIP == "48") |> 
  group_by(SEX, LIT) |> 
  summarise(num = n()) |> 
  show_query()
```

If an analyst wants the result of a database operation to be brought into R as an R object rather than as a database view, the [`collect()`](https://dplyr.tidyverse.org/reference/compute.html) function can be used at the end of a pipeline to load data directly. A companion function from **ipumsr**, [`ipums_collect()`](https://rdrr.io/pkg/ipumsr/man/ipums_collect.html), will add variable and value labels to the collected data based on an IPUMS codebook.

```{r}
literacy_props <- census1910 |> 
  filter(AGE > 18) |> 
  group_by(STATEFIP, SEX, LIT) |> 
  summarise(num = n()) |> 
  group_by(STATEFIP, SEX) |> 
  mutate(total = sum(num, na.rm = T)) |> 
  ungroup() |> 
  mutate(prop = num / total) |> 
  filter(LIT == 4) |> 
  collect()

state_names <- tigris::fips_codes |> 
  select(state_code, state_name) |> 
  distinct()


literacy_props_with_name <- literacy_props %>%
  mutate(STATEFIP = str_pad(STATEFIP, 2, "left", "0")) %>%
  left_join(state_names, by = c("STATEFIP" = "state_code")) %>%
  mutate(sex = ifelse(SEX == 1, "Male", "Female")) 

```

```{r fig.width=10, fig.height=8}
ggplot(literacy_props_with_name,
       aes(x = prop, y = reorder(state_name, prop),
           color = sex)) +
  geom_line(aes(group = state_name), color = "gray10") +
  geom_point(size = 2.5) +
  theme_minimal() +
  scale_color_manual(values = c(Male = "navy", Female = "darkred")) +
  scale_x_continuous(labels = scales::percent) +
  labs(x = "Percent fully literate, 1910",
       color = "",
       y = "")
```
