---
title: "gt Walkthrough"
format: html
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, paged.print = FALSE)
```

```{r}
#| message: false

library(gt)
library(tidyverse)
```

```{r}
data(islands)
```

```{r}
names(islands)
```

```{r}
islands_tbl <- 
  tibble(
    name = names(islands),
    size = islands
  ) |> 
  slice_max(size, n = 10)

islands_tbl
```

```{r}
gt_tbl <- gt(islands_tbl)
gt_tbl
```

# Headers

```{r}
gt_tbl <- 
  gt_tbl |> 
  tab_header(
    title = "Large Landmasses of the World",
    subtitle = "The top ten largest"
  )
gt_tbl
```

> With markdown

```{r}
gt(islands_tbl[1:2, ]) |> 
  tab_header(
    title = md("**Large Landmasses of the Worldd**"),
    subtitle = md("The *top two* largest")
  )
```

# Source Note

```{r}
gt_tbl <- 
  gt_tbl |> 
  tab_source_note(
    source_note = "Source: The World Almanac and Book of Facts, 1975, page 406."
  ) |>
  tab_source_note(
    source_note = md("Reference: McNeil, D. R. (1977) *Interactive Data Analysis*. Wiley.")
  )
gt_tbl
```
# Footnotes

The helper function cells_body() can be used with the location argument to specify which data cells should be the target of the footnote. cells_body() has the two arguments columns and rows. For each of these, we can supply 

- a vector of colnames or rownames
- a vector of column/row indices
- bare column names wrapped in c() or row labels within c(), or 
- a select helper function (starts_with(), ends_with(), contains(), matches(), all_of(), and everything()). For rows specifically, we can use a conditional statement with column names as variables (e.g., size > 15000)

```{r}
gt_tbl <- 
  gt_tbl |> 
  tab_footnote(
    footnote = "The Americas.",
    locations = cells_body(columns = name, rows = 3:4)
  )

gt_tbl
```

```{r}
largest <- islands_tbl |> 
  slice_max(size, n = 1) |> 
  pull(name)

gt_tbl <- 
  gt_tbl |> 
  tab_footnote(
    footnote = md("The *largest* by area."),
    locations = cells_body(
      columns = size,
      rows = name == largest
    )
  ) |> 
  tab_footnote(
    footnote = "The lowest by area",
    locations = cells_body(
      columns = size,
      rows = size == min(size)
    )
  )
gt_tbl
```

# The Stub

Indicate which column is the "stub"

```{r}
gt_tbl <- 
  islands_tbl |> 
  gt(rowname_col = "name")

gt_tbl
```

```{r}
gt_tbl <-
  gt_tbl |>
  tab_stubhead(label = "landmass")
gt_tbl
```

```{r}
# Display the `islands_tbl` data with a stub,
# a heading, source notes, and footnotes
gt_tbl <-
  gt_tbl |>
  tab_header(
    title = "Large Landmasses of the World",
    subtitle = "The top ten largest are presented"
  ) |>
  tab_source_note(
    source_note = "Source: The World Almanac and Book of Facts, 1975, page 406."
  ) |>
  tab_source_note(
    source_note = md("Reference: McNeil, D. R. (1977) *Interactive Data Analysis*. Wiley.")
  ) |>
  tab_footnote(
    footnote = md("The **largest** by area."),
    locations = cells_body(
      columns = size, rows = largest
    )
  ) |>
  tab_footnote(
    footnote = "The lowest by population.",
    locations = cells_body(
      columns = size, rows = contains("arc")
    )
  )

# Show the gt table
gt_tbl
```

# Row groups

```{r}
gt_tbl
```

```{r}
# Create three row groups with the
# `tab_row_group()` function
gt_tbl <-
  gt_tbl |>
  tab_row_group(
    label = "continent",
    rows = 1:6
  ) |>
  tab_row_group(
    label = "country",
    rows = c("Australia", "Greenland")
  ) |>
  tab_row_group(
    label = "subregion",
    rows = c("New Guinea", "Borneo")
  )

# Show the gt table
gt_tbl
```

# Column labels (spans)

```{r}
data(airquality)

airquality_m <- 
  airquality |> 
  mutate(Year = 1973L) |> 
  slice(1:10)

gt_tbl <- 
  gt(airquality_m) |> 
  tab_header(
    title = "New York Air Quality Measurements",
    subtitle = "Daily measurements in New York City (May 1-10, 1973)"
  ) |> 
  tab_spanner(
    label = "Time",
    columns = c(Year, Month, Day)
  ) |> 
  tab_spanner(
    label = "Measurement",
    columns = c(Ozone, Solar.R, Wind, Temp)
  )

gt_tbl
```

> Moving and relabling columns

```{r}
gt_tbl <- 
  gt_tbl |> 
  cols_move_to_start(
    columns = c(Year, Month, Day)
  ) |> 
  cols_label(
    Ozone = html("Ozone,<br>ppbV"),
    Solar.R = html("Solar R.,<br>cal/m<sup>2</sup>"),
    Wind = html("Wind,<br>mph"),
    Temp = html("Temp,<br>&deg;F")
  )
gt_tbl
```
