---
title: "7.4 Distance and proximity analysis"
output: 
  md_document:
    variant: gfm
always_allow_html: true
date: "2025-03-28"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, paged.print = FALSE)
```

# Distance and proximity analysis

```{r}
library(tidycensus)
library(tidyverse)
library(tigris)
library(sf)
options(tigris_use_cache = T)
```

```{r}
nm_tracts <- tracts("NM", cb = TRUE, year = 2019) %>%
  st_transform(32113)
```

```{r}
library(osmdata)
available_tags("amenity")
nm_bb <- getbb("New Mexico")
```

```{r}
nm_hospitals <- nm_bb |> 
  opq() |> 
  add_osm_feature(key = "amenity", value = "hospital") |> 
  osmdata_sf()

nm_hospitals <- nm_hospitals$osm_polygons |> 
  st_transform(32113)

trauma <- nm_hospitals |> 
  filter(str_detect(emergency, "yes")) |> 
  st_cast(to = 'POINT')
```

## Distances

```{r}
nm_trauma <-  trauma |> 
  st_filter(nm_tracts,
            .predicate = st_is_within_distance,
            dist = 100000)
```

```{r}
ggplot() +
  geom_sf(data = nm_tracts, color = "NA", fill = "grey50") +
  geom_sf(data = nm_trauma, color = "red") +
  theme_void()
```

```{r}
dist <- nm_tracts |> 
  st_centroid() |> 
  st_distance(nm_trauma)

dist[1:5, 1:5]
```

The dense matrix shows distances (in meters) between the first five Census
tracts in the dataset and the first five hospitals. When considering
*accessibility*, we may be interested in the distance to the *nearest*
hospital to each Census tract.

```{r}
min_dist <- dist |> 
  apply(1, min) |> 
  as.vector() |> 
  magrittr::divide_by(1000)

hist(min_dist)
```

```{r}
plot(nm_trauma['osm_id'])
ggplot(nm_trauma) +
  geom_sf()
  
```

```{r}
pnt <- nm_trauma$geometry[1,]
class(pnt)
st_coordinates(pnt)[,1]
```

## Travel times

```{r}
library(mapboxapi)
```
```{r eval=FALSE}
times <- mb_matrix(nm_tracts, nm_trauma,
                   allow_large_matrix = TRUE)
```
## Catchment areas with buffers and isochrones

```{r}
ggplot(nm_trauma) +
  geom_sf(aes(color = osm_id)) +
  geom_text(label = nm_trauma$osm_id, 
             x = st_coordinates(nm_trauma$geometry)[,1], 
             y = st_coordinates(nm_trauma$geometry)[,2])
```
```{r}
guadalupe_county <- nm_trauma[nm_trauma$osm_id == "458183369",][1,]

buf5km <- st_buffer(guadalupe_county, dist = 5000)
```

```{r}
iso15min <- mb_isochrone(
  guadalupe_county,
  time = 15,
  profile = "driving-traffic",
  depart_at = "2025-04-19T10:00"
)
```
```{r}
iso15min_p <- st_transform(iso15min, 32113)
buf5km_p <- st_transform(buf5km, 32113)
```


```{r}
library(leaflet)
library(leafsync)
```
```{r}
hospital_icon <- makeAwesomeIcon(icon = "ios-medical",
                                 markerColor = "red",
                                 library = "ion")

m2 <- leaflet() |> 
  addProviderTiles(providers$Esri.WorldStreetMap) |> 
  addPolygons(data = iso15min) |> 
  addAwesomeMarkers(data = st_transform(guadalupe_county, 4326),
                    icon = hospital_icon)
  
m1 <- leaflet() |> 
  addProviderTiles(providers$Esri.WorldStreetMap) |> 
  addPolygons(data = st_transform(buf5km, 4326)) |> 
  addAwesomeMarkers(data =  st_transform(guadalupe_county, 4326),
                    icon = hospital_icon)

sync(m1, m2)
```


## Demographic estimates for zones with areal interpolation


```{r}
guadalupe_poverty <- get_acs(
  geography = "block group",
  variables = c(poverty_denom = "B17010_001",
                poverty_num = "B17010_002"),
  state = "NM",
  county = "Guadalupe",
  geometry = T,
  output = "wide",
  year = 2020
) |> 
  select(poverty_denomE, poverty_numE) |> 
  st_transform(32113)
```

We can then use population-weighted areal interpolation with interpolate_pw() function in tidycensus to estimate family poverty in both the buffer zone and the isochrone. Block weights for Polk County are obtained with tigris, and both the numerator and denominator columns are interpolated.

```{r}
library(glue)

guadalupe_blocks <- blocks(
  state = "NM",
  county = "Guadalupe",
  year = 2020
)


buffer_pov <- interpolate_pw(
  from = guadalupe_poverty,
  to = buf5km,
  extensive = T,
  weights = guadalupe_blocks,
  weight_column = "POP20",
  crs = 32113
) |> 
  mutate(pct_poverty = 100 * (poverty_numE / poverty_denomE))

iso_pov <- interpolate_pw(
  from = guadalupe_poverty,
  to = iso15min,
  extensive = T,
  weights = guadalupe_blocks,
  weight_column = "POP20",
  crs = 32113
) |> 
  mutate(pct_poverty = 100 * (poverty_numE / poverty_denomE))

```

```{r}
data.frame(
  "Method" = c("5km buffer", "15 min isochrome"),
  "Families in poverty" = c(buffer_pov[1,]$poverty_numE, 
                            iso_pov[1,]$poverty_numE),
  "Total families" = c(buffer_pov[1,]$poverty_denomE, 
                       iso_pov[1,]$poverty_denomE),
  "Percent" = c(buffer_pov[1,]$pct_poverty, 
                iso_pov[1,]$pct_poverty)
)
```


