---
title: "8.1-Segregation and Diversity Indices"
output: 
  md_document:
    variant: gfm
always_allow_html: true
date: "2025-04-18"
---

# Modeling segregation and diversity

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, paged.print = FALSE)
```

```{r}
library(tidycensus)
library(tidyverse)
library(segregation)
library(tigris)
library(sf)
```

## Data Preparation

```{r}
nm_acs_data <- get_acs(
  geography = "tract",
  variables = c(
    white = "B03002_003",
    black = "B03002_004",
    native = "B03002_005",
    asian = "B03002_006",
    hispanic = "B03002_012"
  ),
  state = "NM",
  geometry = T,
  year = 2019
)
```

```{r}
us_urban_areas <- get_acs(
  geography = "urban area",
  variables = "B01001_001",
  geometry = T,
  year = 2019,
  survey = "acs1"
) |> 
  filter(estimate >= 500000) |> 
  transmute(urban_name = str_remove(NAME,
                                    fixed(", NM Urbanized Area (2010)")))
```

> Join tracts and urbanized areas

```{r}
nm_urban_data <- nm_acs_data |> 
  st_join(us_urban_areas, left = FALSE) |> 
  select(-NAME) |> 
  st_drop_geometry()
```

```{r}
head(nm_urban_data)
```

## Dissimilarity index

$$
D = \frac{1}{2}\sum_{i=1}^N\left| \frac{a_i}{A}-\frac{b_i}{B} \right|
$$

where ai represents the population of group A in a given areal unit i; A is
the total population of that group in the study region (e.g. a metropolitan
area); and bi and B are the equivalent metrics for the second group. The
index ranges from a low of 0 to a high of 1, where 0 represents perfect
integration between the two groups and 1 represents complete segregation.
This index is implemented in the **segregation** package with the
[`dissimilarity()`](https://elbersb.github.io/segregation/reference/dissimilarity.html)
function.

> Dissimilarity White and Hispanic in Albuquerque

```{r}
nm_urban_data <- nm_urban_data |> 
  mutate(urban_name = str_remove(urban_name,
                                    fixed(", TX--NM Urbanized Area (2010)")))
```

```{r}
nm_urban_data |> 
  filter(variable %in% c("white", "hispanic"),
         urban_name == "Albuquerque") |> 
  dissimilarity(
    group = "variable",
    unit = "GEOID",
    weight = "estimate"
  )
```

> The same exercise for SF-Oakland gives D = 0.5135526

```{r}
nm_urban_data |> 
  filter(variable %in% c("white", "hispanic")) |>
  group_by(urban_name) |> 
  group_modify(~
    dissimilarity(.x,
      group = "variable",
      unit = "GEOID",
      weight = "estimate"
    )
  ) |> 
  arrange(desc(est))
```

> El Paso is much more segregated than Albuquerque

## Multigroup segregation indices

Mutual Information Index $M$:

$$
M(\text{T}) = \sum_{u=1}^U \sum_{g=1}^G p_{ug}\log\frac{p_{ug}}{p_u p_g}
$$

where U is the total number of units u, G is the total number of groups g,
and pug is the joint probability of being in unit u and group g, with pu and
pg referring to unit and group probabilities.

Theil's H:

$$
H(\text{T}) = \frac{M(\text{T})}{E(\text{T})}
$$

where E(T) is the entropy of T, normalizing H to range between values of 0
and 1.

```{r}
mutual_within(
  data = nm_urban_data,
  group = "variable",
  unit = "GEOID",
  weight = "estimate",
  within = "urban_name",
  wide = T
)
```

> Why are these results seemingly opposite of the single-group comparison?
>
> `mutual_local` breaks down the M into unit-level scores

```{r}
abq_local_seg <-  nm_urban_data |> 
  filter(urban_name == "Albuquerque") |> 
  mutual_local(
    group = "variable",
    unit = "GEOID",
    weight = "estimate",
    wide = T
  )
head(abq_local_seg)
```

```{r}
abq_tracts_seg <- tracts("NM", cb = T, year = 2019) |> 
  inner_join(abq_local_seg, by = "GEOID")
```

```{r}
abq_tracts_seg |> 
  ggplot(aes(fill = ls)) +
  geom_sf(color = NA) +
  coord_sf(crs = 32113) +
  scale_fill_viridis_c(option = "inferno") +
  theme_void() +
  labs(fill = "Local\nsegregation\nindex")
```

## Diversity gradient

Expectation is that diversity decreases with distance from urban center. The
entropy index is

$$
E=\sum_{r=1}^n Q_r \ln\frac{1}{Q_r}
$$

where $Q_r$ is the proportion of group $r$'s proportion of the population in
the geographic unit.

```{r}
abq_entropy <- nm_urban_data |> 
  filter(urban_name == "Albuquerque") |> 
  group_by(GEOID) |> 
  group_modify(~data.frame(entropy = entropy(
    data = .x,
    group = "variable",
    weight = "estimate",
    base = 5 # number of groups in the calculation
  )))

abq_entropy_geo <- tracts("NM", cb = T, year = 2019) |> 
  inner_join(abq_entropy, by = "GEOID")
```

> get driving times

```{r}
library(mapboxapi)

abq_city_hall <- mb_geocode("City Hall, Albuquerque NM")

minutes_to_downtown <- mb_matrix(abq_entropy_geo, abq_city_hall)
```

```{r}
abq_entropy_geo$minutes <- as.numeric(minutes_to_downtown)

ggplot(abq_entropy_geo, aes(x = minutes_to_downtown, y = entropy)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "loess") +
  theme_minimal() +
  scale_x_continuous(limits = c(0, 80)) +
  labs(title = "Diversity gradient, ABQ urbanized area",
       x = "Travel time to downtown in minutes",
       y = "Entropy index")
```

There seems to be no relationship to diversity and travel time.!
