---
title: "08.4-Geographically Weighted Regression"
output: 
  md_document:
    variant: gfm
date: "2025-04-19"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, paged.print = FALSE)
```

# Geographically weighted regression

A way to look at local variation from neighborhood to neighborhood in the
relationship between predictors and outcomes. It uses a kernel weighting
function (distance-decay).

$$
Y_i=\alpha_i+\sum^m_{k=1}\beta_{ik}X_{ik}+\epsilon_i
$$

where the model intercept, parameters, and error term are all
location-specific. Notably, $\beta_{ik}$ represents a *local regression
coefficient* for predictor $k$ (of the total number of predictors $m$) that
is specific to location $i$.

## Prepare data

```{r}
library(tidycensus)
library(tidyverse)
library(sf)
```

```{r}
nm_counties <- c("Valencia", "Bernalillo","Sandoval", 
                 "Santa Fe", "Los Alamos")

variables_to_get <- c(
  median_value = "B25077_001",
  median_rooms = "B25018_001",
  median_income = "DP03_0062",
  total_population = "B01003_001",
  median_age = "B01002_001",
  pct_college = "DP02_0068P",
  pct_foreign_born = "DP02_0094P",
  pct_white = "DP05_0077P",
  median_year_built = "B25037_001",
  percent_ooh = "DP04_0046P"
)

abq_data <- get_acs(
  geography = "tract", 
  variables = variables_to_get,
  state = "NM",
  geometry = T,
  county = nm_counties,
  output = "wide",
  year = 2020
) |> 
  select(-NAME) |> 
  st_transform(32113)

library(units)

abq_data_for_model <- abq_data |> 
  mutate(pop_density = as.numeric(set_units(total_populationE / st_area(geometry), "1/km2")),
         median_structure_age = 2018 - median_year_builtE)  |> 
    select(!ends_with("M")) |> 
  rename_with(.fn = ~str_remove(.x, "E$")) |> 
  na.omit()

formula1 <- "log(median_value) ~ median_rooms + median_income + pct_college + pct_foreign_born + pct_white + median_age + median_structure_age + percent_ooh + pop_density + total_population"

model1 <- lm(formula = formula1, data = abq_data_for_model)

summary(model1)
```

## Choosing a bandwidth

> -   Fixed: uses a cutoff distance
> -   Adaptive: uses nearest neighbors
>
> Census tracts vary widely in size, so adaptive kernels are usually used.
>
> GWmodel can help choose, but object must be converted from `sf` to
> `SpatialPolygonsDataFrame`.

```{r}
library(GWmodel)
library(sf)
```

```{r}
bw <- bw.gwr(
  formula = formula1,
  data = abq_data_for_model,
  kernel = "bisquare",
  adaptive = TRUE
) 
```

```{r}
bw
```

## Fitting and evaluating the GWR model

```{r}
gw_model <- gwr.basic(
  formula = formula1,
  data = abq_data_for_model,
  bw = bw,
  kernel = "bisquare", 
  adaptive = TRUE
)
gw_model
```

```{r}
names(gw_model)
```

> `SDF` is a `SpatialPolygonsDataFrame` that can be mapped.

```{r}
gw_model_results <- gw_model$SDF |> 
  st_as_sf()

names(gw_model_results)
```

> Map the local R-squared

```{r}
ggplot(gw_model_results, aes(fill = Local_R2)) +
  geom_sf(color = NA) +
  scale_fill_viridis_c() +
  theme_void()
  
```

> The model performs better in the Santa Fe and Los Alamos regions and worse
> in Bernalillo County and near Albuquerque.

```{r eval=FALSE}
library(mapview)
mapview(gw_model_results, zcol = "Local_R2")
```

> Visualize local parameter estimate variation

```{r}
ggplot(gw_model_results, aes(fill = pct_college)) +
  geom_sf(color = NA) + 
  scale_fill_viridis_c() + 
  theme_void() + 
  labs(fill = "Local β for \npct_college")  
```

```{r}
ggplot(gw_model_results, aes(fill = pct_white)) +
  geom_sf(color = NA) + 
  scale_fill_viridis_c() + 
  theme_void() + 
  labs(fill = "Local β for \npct_white") 
```

```{r}
ggplot(gw_model_results, aes(fill = median_structure_age)) +
  geom_sf(color = NA) + 
  scale_fill_viridis_c() + 
  theme_void() + 
  labs(fill = "Local β for \nmedian_structure_age") 
```
