---
title: "10-AnalyzingCensusMicrodata"
output: 
  md_document:
    variant: gfm
date: "2025-04-19"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, paged.print = FALSE)
```

# Analyzing Census Microdata

## PUMS and tidyverse

### Basic weight tabulation

```{r}
library(tidycensus)
library(tidyverse)

nm_pums <- get_pums(
  variables = c("SEX", "AGEP"),
  state = "NM",
  survey = "acs5",
  year = 2023,
  recode = T
)
nm_pums
```

```{r}
sum(nm_pums$PWGTP)
```

```{r}
nm_pums |> count(wt = PWGTP)
```

> Tabulate by age and sex

```{r}
nm_pums |> 
  count(SEX_label, AGEP, wt = PWGTP)
```

> Number of people over 65

```{r}
nm_pums |> 
  filter(AGEP >= 65) |> 
  count(SEX, wt = PWGTP)
```

> compare with `get_acs`

```{r}
get_acs(geography = "state",
        state = "NM",
        variables = c("DP05_0030", "DP05_0031"),
        year = 2023)
```

> Because PUMS data consist of a subset of the full ACS sample, tabulations from the ACS PUMS will not match those from published tables of ACS data.

### Group-wise tabulation

> How does rent burden vary by race/ethnicity and household type?

This requires obtaining data on rent burden (gross rent as percentage of household income) with variable `GRPIP`; race and ethnicity with variables `RAC1P` and `HISP`; and household type with variable `HHT`.

```{r}
hh_variables <- c("PUMA", "GRPIP", "RAC1P", 
                  "HISP", "HHT")

nm_hh_data <- get_pums(
  variables = hh_variables,
  state = "NM",
  year = 2023,
  variables_filter = list(
    SPORDER = 1,
    TEN = 3  # Households paying cash rent
  )
)
nm_hh_data
```

```{r}
nm_hh_recoded <- nm_hh_data |> 
  mutate(
    race_ethnicity = case_when(
      HISP != "01" ~ "Hispanic",
      HISP == "01" & RAC1P == "1" ~ "White",
      HISP == "01" & RAC1P == "2" ~ "Black",
      HISP == "01" & RAC1P == "3" ~ "Native",
      TRUE ~ "Other"
    ),
    married = case_when(
      HHT == "1" ~ "Married",
      TRUE ~ "Not married"
    )
  )

nm_hh_summary <- nm_hh_recoded |> 
  filter(race_ethnicity != "Other") |> 
  group_by(race_ethnicity, married) |> 
  summarize(
    prop_above_40 = sum(WGTP[GRPIP >= 40]) / sum(WGTP)
  )

nm_hh_summary
```

> 43.5% of unmarried black households pay over 40% of their incomes in gross rent

## Mapping PUMS data

```{r}
library(tigris)
library(tmap)
options(tigris_use_cache = T)

nm_pumas <- pumas("NM", year = 2020)

plot(nm_pumas$geometry)
```

> prepare data for map

```{r}
nm_data_for_map <- nm_hh_recoded |> 
  group_by(race_ethnicity, married, PUMA) |> 
  summarise(
    percent_above_40 = 100 * (sum(WGTP[GRPIP >= 40]) / sum(WGTP))
  ) |> 
  filter(race_ethnicity == "Native",
         married == "Not married")
```

```{r message=FALSE}
library(tmap)

joined_pumas <- nm_pumas |> 
  left_join(nm_data_for_map, by = c("PUMACE10" =  "PUMA"))

tm_shape(joined_pumas) +
  tm_polygons(col = "percent_above_40",
              palette = "Reds",
              title = "% rent-burdened\nunmarried Native households") +
  tm_layout(legend.outside = T,
            legend.outside.posistion = "right")
```

## Survey design and ACS PUMS

### Get data with replicate weights

In order to calculate std. errors and moes, a special formula must be used involving *replicate weights* for each observation and *Successive Difference Replication*.

> Download data with household replicate weights.
>
> Note that TEN is included, filtering must take place after the calculations are performed to ensure that they take into account the full structure.

```{r}
nm_hh_replicate <- get_pums(
  variables = c("TEN", hh_variables),
  state = "NM",
  recode = T,
  year = 2023,
  variables_filter = list(
    SPORDER = 1
  ),
  rep_weights = "housing"
)
```

```{r}
names(nm_hh_replicate)
```

### Survey object

```{r}
library(survey)
library(srvyr)

nm_hh_svy <- nm_hh_replicate |> 
  to_survey(type = "housing",
            design = "rep_weights") |> 
  filter(TEN == 3)

class(nm_hh_svy)
```

### Estimates and errors

```{r}
nm_hh_svy |> 
  survey_count(PUMA, HHT_label) |> 
  head()
```

> the survey object has its own functions for count and mean, for example.

```{r}
nm_svy_summary <- nm_hh_svy |> 
  mutate(
    race_ethnicity = case_when(
      HISP != "01" ~ "Hispanic",
      HISP == "01" & RAC1P == "1" ~ "White",
      HISP == "01" & RAC1P == "2" ~ "Black",
      HISP == "01" & RAC1P == "3" ~ "Native",
      TRUE ~ "Other"
    ),
    married = case_when(
      HHT == "1" ~ "Married",
      TRUE ~ "Not married"
    ),
    above_40 = GRPIP >= 40
  ) |> 
  filter(race_ethnicity != "Other") |> 
  group_by(race_ethnicity, married) |> 
  summarize(
    prop_above_40 = survey_mean(above_40)
  )
head(nm_svy_summary)
```

### Convert standard errors to margins of error.

To convert standard errors to *margins of error* around the derived PUMS estimates, analysts should multiply the standard errors by the following coefficients:

-   90 percent confidence level: 1.645
-   95 percent confidence level: 1.96
-   99 percent confidence level: 2.576

```{r}
nm_svy_summary_moe <- nm_svy_summary |> 
  mutate(prop_above_40_moe = prop_above_40_se * 1.645,
         label = paste(race_ethnicity, married, sep = ", "))


nm_svy_summary_moe |> 
  ggplot(aes(x = prop_above_40,
             y = reorder(label,
                         prop_above_40))) +
  geom_errorbar(aes(xmin = prop_above_40 - prop_above_40_moe,
                    xmax = prop_above_40 + prop_above_40_moe)) +
  geom_point(size = 3, color = "navy") +
  labs(title = "Rent burdened-households in New Mexico",
       x = "2018-2023 ACS estimate (from PUMS data)",
       y = "",
       caption = "Rent-burdened defined when gross rent is 40 percent or more\nof household income. Error bars represent a 90 percent confidence level.") +
  scale_x_continuous(labels = scales::percent) +
  theme_grey(base_size = 12)
  
```

## Modeling PUMS data

```{r}
nm_pums_to_model <- get_pums(
    variables = c("PUMA", "SEX", "MIG",
                "AGEP", "SCHL", "WAGP", 
                "COW", "ESR", "MAR", "NOC"),
  state = "NM",
  survey = "acs5",
  year = 2023,
  rep_weights = "person"
)
```

### Data prep

The target variable is coded 0 or 1 since we will estimate a logistic regression model with a binary outcome.

```{r}
nm_pums_recoded <- nm_pums_to_model %>%
  mutate(
    emp_type = case_when(
      COW %in% c("1", "2") ~ "private",
      COW %in% c("3", "4", "5") ~ "public",
      TRUE ~ "self"
      ), 
    child = case_when(
      NOC > 0 ~ "yes",
      TRUE ~ "no"
    ),
    married = case_when(
      MAR == 1 ~ "yes",
      TRUE ~ "no"
    ),
    college = case_when(
      SCHL %in% as.character(21:24) ~ "yes",
      TRUE ~ "no"
    ),
    sex = case_when(
      SEX == 2 ~ "female",
      TRUE ~ "male"
    ),
    migrated = case_when(
      MIG == 1 ~ 0,
      TRUE ~ 1
    )
  )
```

> Focus on individuals aged 25-49 who are employed and have earned wages in the past year.

```{r}
nm_model_svy <- nm_pums_recoded %>% 
  to_survey() %>%
  filter(
    ESR == 1,   # civilian employed
    WAGP > 0,   # earned wages last year
    AGEP >= 25,
    AGEP <= 49
  ) %>%
  rename(age = AGEP, wages = WAGP)
```

### Fit and evaluate

```{r}
migration_model <- svyglm(
  formula = migrated ~ log(wages) + sex + age + emp_type + 
    child + married + college + PUMA,
  design = nm_model_svy,
  family = quasibinomial()
) 
```

```{r}
summary(migration_model)
```
