---
title: "10-AnalyzingCensusMicrodata"
output: 
  md_document:
    variant: gfm
date: "2025-04-19"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, paged.print = FALSE)
```

# NHGIS historical data

Data citation

Steven Manson, Jonathan Schroeder, David Van Riper, Katherine Knowles, Tracy Kugler, Finn Roberts, and Steven Ruggles. IPUMS National Historical Geographic Information System: Version 19.0 [dataset]. Minneapolis, MN: IPUMS. 2024. <http://doi.org/10.18128/D050.V19.0>

The example below uses `read_nhgis_sf()` to read in spatial and demographic data on immigrants in New York City in 1910. As the 1910 shapefile folder includes both NYC Census tracts and a separate dataset with US counties, the top-level folder should be unzipped, `shape_file` pointed to the second-level zipped folder, and the `shape_layer` argument used to exclusively read in the tracts. The [`filter()`](https://dplyr.tidyverse.org/reference/filter.html) call will drop Census tracts that do not have corresponding data (so, outside NYC).

```{r}
library(ipumsr)
```
