---
title: "07.5-Spatial Neighborhoods and weights"
output: 
  md_document:
    variant: gfm
always_allow_html: true
date: "2025-04-18"
---

# Spatial Neighborhoods and weights matrices

ESDA with spatial properties. Example, median age by census tract.

```{r}
library(tidycensus)
library(tidyverse)
library(tigris)
library(sf)
library(spdep)
options(tigris_use_cache = T)
```

```{r}
abq <- core_based_statistical_areas(
  cb = T,
  year = 2020) |> filter(str_detect(NAME, "Albuquerque")) |> 
  st_transform(32113)
  
abq_tracts <- get_acs(
  geography = "tract",
  variables = "B01002_001",
  state = "NM",
  year = 2020,
  geometry = T
) |> 
  st_transform(32113) |> 
  st_filter(abq, .predicate = st_within) |> 
  na.omit()

ggplot(abq_tracts) +
  geom_sf(aes(fill = estimate), color = NA) +
  scale_fill_viridis_c() +
  theme_void()
```

## Spatial neighborhood definitions

`spdep` includes:

-   *Proximity-based neighbors*, where neighboring features are
    identified based on some measure of distance. Neighbors might be defined
    as those that fall within a given distance threshold (e.g. all features
    within 2km of a given feature) or as *k*-nearest neighbors (e.g. the
    nearest eight features to a given feature).
-   *Graph-based neighbors*, where neighbors are defined through network
    relationships (e.g. along a street network).
-   *Contiguity-based neighbors*, used when geographic features are
    polygons. Options for contiguity-based spatial relationships include
    *queen’s case neighbors*, where all polygons that share at least one
    vertex are considered neighbors; and *rook’s case neighbors*, where
    polygons must share at least one line segment to be considered
    neighbors.

## Queen contiguity

```{r}
neighbors <- poly2nb(abq_tracts, queen = T)

summary(neighbors)
```

```{r}
abq_coords <- abq_tracts |> 
  st_centroid() |> 
  st_coordinates()

```

```{r}
plot(abq_tracts$geometry)
plot(neighbors, 
     coords = abq_coords,
     add = T,
     col = "blue", points = F)
```

```{r}
neighbors[[1]]
```

## Generate spatial weights matrix

> Convert neighbors list to spatial weights
>
> Can use row-standardized `style = "W", when analyzing means` or 1/0 with
> `style = "B" when computing sums`.

```{r}
weights <- nb2listw(neighbors, style = "W")
weights$weights[[1]]
```

## Global and local spatial autocorrelation

The degree to which a variable is similar to or different from their
neighbors.

-   *Spatial clustering*: data values tend to be similar to neighboring data
    values;
-   *Spatial uniformity*: data values tend to differ from neighboring data
    values;
-   *Spatial randomness*: there is no apparent relationship between data
    values and those of their neighbors.

### Spatial lags and Moran's $I$

```{r}
abq_tracts$lag_estimate <- lag.listw(weights, abq_tracts$estimate)
```

```{r}
ggplot(abq_tracts, aes(x = estimate, y = lag_estimate)) +
  geom_point(alpha = 0.3) +
  geom_abline(color = "red") +
  theme_minimal() +
  labs(title = "Median age by Census tract, Albuquerque Metro",
       x = "Median age",
       y = "Spatial lag, median age", 
       caption = "Data source: 2016-2020 ACS via the tidycensus R package.\nSpatial relationships based on queens-case polygon contiguity.")
```

> positive correlation suggested, quantify with Moran's $I$

```{r}
moran.test(abq_tracts$estimate, weights)
```

> Positive statistic shows age populations are spatially clustered, not
> spatially uniform.

## Local spatial autocorrelation

> Getis-Ord local G

```{r}
localg_weights <- nb2listw(include.self(neighbors))

abq_tracts$localG <- localG(abq_tracts$estimate, localg_weights)

ggplot(abq_tracts) + 
  geom_sf(aes(fill = as.numeric(localG)), color = NA) + 
  scale_fill_distiller(palette = "RdYlBu") + 
  theme_void() + 
  labs(fill = "Local Gi* statistic")
```

> Hotspots

```{r}
abq_tracts <- abq_tracts |> 
  mutate(hotspot = case_when(
    localG >= 2.576 ~ "High cluster",
    localG <= -2.576 ~ "Low cluster",
    TRUE ~ "Not significant"
  ))

ggplot(abq_tracts) +
  geom_sf(aes(fill = hotspot), color = "grey90", size = 0.1) +
  scale_fill_manual(values = c("red", "blue", "grey")) +
  theme_void()
```

```{r}
class(abq_tracts$localG)
```

```{r}
abq_tracts[2:10,]$localG
```

## Local indicators of spatial association (LISA)

One of the major benefits of using LISA for exploratory analysis is its
ability to identify both *spatial clusters*, where observations are
surrounded by similar values, and *spatial outliers*, where observations are
surrounded by dissimilar values.

```{r}
set.seed(1916)

abq_tracts$scaled_estimate <- as.numeric(scale(abq_tracts$estimate))

abq_lisa <- localmoran_perm(
  abq_tracts$scaled_estimate,
  weights,
  nsim = 999L,
  alternative = "two.sided"
) |> 
  as_tibble() |> 
  set_names(c("local_i", "exp_i", "var_i", "z_i", "p_i",
              "p_i_sim", "pi_sim_folded", "skewness", "kurtosis"))

abq_lisa_df <- abq_tracts |> 
  select(GEOID, scaled_estimate) |> 
  mutate(lagged_estimate = lag.listw(weights, scaled_estimate)) |> 
  bind_cols(abq_lisa)

abq_lisa_df
```

```{r}
abq_lisa_clusters <- abq_lisa_df %>%
  mutate(lisa_cluster = case_when(
    p_i >= 0.05 ~ "Not significant",
    scaled_estimate > 0 & local_i > 0 ~ "High-high",
    scaled_estimate > 0 & local_i < 0 ~ "High-low",
    scaled_estimate < 0 & local_i > 0 ~ "Low-low",
    scaled_estimate < 0 & local_i < 0 ~ "Low-high"
  ))
```

```{r}
color_values <- c(`High-high` = "red", 
                  `High-low` = "pink", 
                  `Low-low` = "blue", 
                  `Low-high` = "lightblue", 
                  `Not significant` = "white")

ggplot(abq_lisa_clusters, aes(x = scaled_estimate,
                              y = lagged_estimate,
                              fill = lisa_cluster)) +
  geom_point(color = "black", shape = 21, size = 2) +
  theme_minimal() +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_vline(xintercept = 0, linetype = "dashed") +
  scale_fill_manual(values = color_values) +
  labs(x = "Median age (z-score)",
       y = "Spatial lag of median age (z-score)",
       fill = "Cluster type")
```

```{r}
ggplot(abq_lisa_clusters, aes(fill = lisa_cluster)) +
  geom_sf(size = 0.1) +
  theme_void() +
  scale_fill_manual(values = color_values) +
  labs(fill = "Cluster type")
```
